A generalization of the Bradley-Terry model for draws in chess

### 2019-05-07 inst/collusion_chess_draws.Rnw
work

%% just as usual
\author{Robin K. S. Hankin\\Auckland University of Technology}
\title{Partial rank data with the \pkg{hyper2} package: likelihood
  functions for generalized Bradley-Terry models}

%\VignetteIndexEntry{The hyper2 package}

@@

In inference problems where the dataset is outcomes of paired
comparisons, such as chess, The Bradley-Terry model can be augmented
to accommodate draws in a number of ways.  Here I present a new
augmentation of Bradley-Terry in which a drawn chess match is regarded
as a three-way competition between the players and an entity that wins
if the match is drawn.  The resulting likelihood functions are easily
optimised numerically and I present a number of results.

@@

\begin{align}\nonumber
\prod_{i<j}&{}
\left(\frac{p_i+p_W}{p_i+p_j+p_D+p_W}\right)^{W_{ij}^\mathrm{white}}
\left(\frac{p_j}{p_i+p_j+p_D+p_W    }\right)^{L_{ij}^\mathrm{white}}\\
&{} \left(\frac{p_i}{p_i+p_j+p_D+p_W}\right)^{W_{ij}^\mathrm{black}}\label{likelihood_kka}
\left(\frac{p_j+p_W}{p_i+p_j+p_D+p_W    }\right)^{L_{ij}^\mathrm{black}}\\
&{} \left(\frac{p_D}{p_i+p_j+p_D}\right)^{D_{ij}}
\nonumber
\end{align}

@@

which might be preferable on grounds of clarity.  We now apply this
likelihood model to real datasets.

@@

Grandmasters Anand, Karpov, Kasparov have long-standing rivalry and
have played many matches amongst themselves.  A summary may be found
in Table~\ref{kka}.

@@

\begin{figure}[htbp]
\begin{center}
<<ddc,echo=TRUE,fig=TRUE>>=
dotchart(maxp(karpov_kasparov_anand),pch=16,xlim=c(0,0.55))
@
\caption{Maximum likelihood estimate for the strengths of the 3 Grandmasters in Table~\ref{kka} 
\label{maxp_karpov_kasparov_anand} under a likelihood defined as Equation~\ref{likelihood_kka}}
  \end{center}
\end{figure}

and this is show visually in figure~\ref{maxp_karpov_kasparov_anand},
which suggests a number of interesting and testable statistical
hypotheses.  Firstly, we observe that $D=0$ cannot possibly be the
case: the likelihood would be identically zero, owing to the nonzero
term in the numerator of the last term of
Equation~\ref{likelihood_kka}.  However, we see that the maximum
likelihood estimate for Draw strength $\hat{D}\simeq 0.53$: compare
the na\"ive estimate of $\frac{224}{365}\simeq 0.61$ given by
observing 224 draws among 265 matches.  The MLE is lower because the
maximization process automatically adjusts for the different players'
strengths.  Secondly, we may make inferences about $W$, the extra
strength conferred by playing White: it is conceivable that $W=0$ as
this term does not appear on its own in the numerator of any term.  We
may test $H_0\colon W=0$ using the method of support by maximizing the
likelihood subject the constraint $W=0$ and comparing with the
likelihood when the optimization is free.

@@

e <- function(white){
    maxp(
        karpov_kasparov_anand,
        startp=c(1/3-small,1/3-small,1/3-small,small),
        fcm=-cbind(0, 0, 0, 1),fcv= -white,give=TRUE)$value
}
e(0.001)
e(0.002)
e(0.003)
maxp(karpov_kasparov_anand,give=TRUE)$value
328.2975-335.6025
@

@@

One very natural hypothesis would be that the players are of equal
strength and we migth write $H_=\colon
p_\mathrm{Kr}=p_\mathrm{Ks}=p_\mathrm{An}$.  This too is readily
testable by the device of calculating the maximum support over the
region in which $p_\mathrm{Kr}=p_\mathrm{Ks}=p_\mathrm{An}$.

<<print=FALSE,echo=FALSE>>=
objective <- function(V){
  x <- V[1]  # strength of all three players
  w <- V[2]  # white player advantage
  D <- 1-3*x-w  # Draw monster
  return(  
      -loglik(karpov_kasparov_anand,c(x,x,x,w))  # negative as we minimize
  )
}

@@

extra_support <-   unconstrained + constrained  # negative used

pchisq(2*extra_support,df=2,lower.tail=FALSE)
@

In this case the extra support in going from the constrained to the
free optimization is about 4.93.  However, this time the constraint
corresponds to a loss of {\em two} degrees of freedom so we would need
4 units of support, which is attained in this case.  Alternatively we
might follow Wilks, and observe that the value of
$2\Lambda=2*4.92=9.84$ lies in the tail region of its asymptotic
distribution with a p-value of about 0.0072.  We can justifiably
assert that the players do indeed have different strengths, even if we
account for white advantage and the proclivity to draw.

@@

>>>>

%% just as usual
\author{Robin K. S. Hankin\\Auckland University of Technology}
\title{A generalization of the Bradley-Terry model for draws in chess}

%\VignetteIndexEntry{chess draws}

@@

In inference problems where the dataset is outcomes of paired
comparisons, such as chess, the Bradley-Terry model can be augmented
to accommodate draws in a number of ways.  Here I present a new
augmentation of Bradley-Terry in which a drawn chess match is regarded
as a three-way competition between the players and an entity that wins
if the match is drawn.  The resulting likelihood functions are easily
optimised numerically and I present a number of results.

@@

\begin{align}\nonumber
\prod_{i<j}&{}
\left(\frac{p_i+p_W}{p_i+p_j+p_D+p_W}\right)^{W_{ij}^\mathrm{white}}
\left(\frac{p_j}{p_i+p_j+p_D+p_W    }\right)^{L_{ij}^\mathrm{white}}\\
&{} \left(\frac{p_i}{p_i+p_j+p_D+p_W}\right)^{W_{ij}^\mathrm{black}}\label{likelihood_WD}
\left(\frac{p_j+p_W}{p_i+p_j+p_D+p_W    }\right)^{L_{ij}^\mathrm{black}}\\
&{} \left(\frac{p_D}{p_i+p_j+p_D}\right)^{D_{ij}}
\nonumber
\end{align}

@@

which might be preferable on grounds of clarity.  We now apply this
likelihood model to real chess datasets.  The general approach to
testing any hypothesis $H$ on any likelihood function $\mathcal{L}$ is
first to identify the constraint on hypothesis space corresponding to
$H$.  One then finds the unconstrained maximum likelihood
$\mathcal{L}_\mathrm{free}$ and compares it with the maximum
likelihood $\mathcal{L}_\mathrm{cons}$ when the optimization is
constrained by $H$.  The test statistic is then the likelihood ratio
$\mathcal{L}_\mathrm{free}/\mathcal{L}_\mathrm{cons}$ or, equivalently
the support $\mathcal{S}=\log\mathcal{R}$.  Wilks's theorem states
that the asymptotic distribution of $2\mathcal{S}$ is $\chi^2_n$ where
$n$ is the number of degrees of freedom implemented by $H$.

@@

Grandmasters Anand, Karpov, Kasparov have long-standing rivalry and
have played many matches amongst themselves.  A summary may be found
in Table~\ref{kka}.  Equation~\ref{likelihood_WD} becomes

@@

\begin{figure}[htbp]
\begin{center}
<<ddc,echo=TRUE,fig=TRUE>>=
dotchart(maxp(karpov_kasparov_anand),pch=16,xlim=c(0,0.55))
@
\caption{Maximum likelihood estimate for the strengths of the 3 Grandmasters in Table~\ref{kka} 
\label{maxp_karpov_kasparov_anand} under a likelihood defined as Equation~\ref{likelihood_WD}}
  \end{center}
\end{figure}

and this is shown visually in figure~\ref{maxp_karpov_kasparov_anand},
which suggests a number of interesting and testable statistical
hypotheses.  Firstly, we observe that $D=0$ cannot possibly be the
case: the likelihood for this hypothesis would be identically zero,
owing to the nonzero term in the numerator of the last term of
Equation~\ref{likelihood_WD}.  However, we see that the maximum
likelihood estimate for Draw strength $\hat{D}\simeq 0.53$: compare
the na\"ive estimate of $\frac{224}{365}\simeq 0.61$ given by
observing 224 draws among 265 matches.  The MLE is lower because the
maximization process automatically adjusts for the different players'
strengths.  Secondly, we may make inferences about $W$, the extra
strength conferred by playing White: it is conceivable that $W=0$ as
this term does not appear on its own in the numerator of any term.  We
may test $H_0\colon W=0$ using the method of support by maximizing the
likelihood subject the constraint $W=0$ and comparing with the
likelihood when the optimization is free.

@@

e <- function(white){
    maxp(
        karpov_kasparov_anand,
        startp=c(1/3-small,1/3-small,1/3-small,small),
        fcm=-cbind(0, 0, 0, 1),fcv= -white,give=TRUE)$value
}
ignore1 <- e(0.001)
ignore2 <- e(0.002)
ignore3 <- e(0.003)
ignore4 <- maxp(karpov_kasparov_anand,give=TRUE)$value
support <- 328.2975-335.6025
@

@@

<<kka_equal_strength, print=FALSE,echo=FALSE>>=
objective <- function(V){
  x <- V[1]  # strength of all three players
  w <- V[2]  # white player advantage
  D <- 1-3*x-w  # Draw monster
  return(  
      -loglik(karpov_kasparov_anand,c(x,x,x,w))  # negative as we minimize
  )
}

@@

extra_support <-   unconstrained + constrained  # negative used
pvalue <- pchisq(2*extra_support,df=2,lower.tail=FALSE)
@

might write $H_=\colon\mathbf{Kr}=\colon\mathbf{Ks}=\colon\mathbf{An}$

One very natural and interesting hypothesis would be that the three
players are of equal strength
%equal_strength_is_assessed_in_chunk_kka_equal_strength_above and we
might write $H_=\colon p_\mathrm{Kr}=p_\mathrm{Ks}=p_\mathrm{An}$.
This too is readily testable by the device of calculating the maximum
support over the region in which
$p_\mathrm{Kr}=p_\mathrm{Ks}=p_\mathrm{An}$.  In this case the extra
support in going from the constrained to the free optimization is
about 4.93.  However, this time the constraint corresponds to a loss
of {\em two} degrees of freedom so we would need 4 units of support,
which is attained in this case.  Alternatively we might follow Wilks,
and observe that the value of $2\mathcal{S}=2*4.92=9.84$ lies in the
tail region of its asymptotic distribution with a p-value of about
0.0072.  We can justifiably assert that the players do indeed have
different strengths, even if we account for white advantage and the
proclivity to draw.

@@

<<echo=FALSE,print=FALSE>>=
do_from_scratch  <- FALSE  # set to TRUE to recalculate (slow, maybe 2 mins)
hotstart <- indep(maxp(collusion))

f <- function(LO,give=FALSE){
    small <- 1e-5

    eq <- hotstart
    m <- sum(eq[2:3])
    a <- exp(LO)
    eq[2:3] <- m/(1+a)*c(1,a)  # exact

    eq[2:3] <- eq[2:3] + c(-small,small)

    out <-
        maxp(collusion,
             startp=eq,
             fcm=c(0, -1, exp(LO),rep(0,22)),
             fcv=0,
             give=TRUE)

    if(give){return(out)}else{return(out$value)}
}

x <- seq(from= 1.5,to = 2.7,len=32)
if(do_from_scratch){
    jj <- sapply(x,f)
} else {
    jj <-
        c(-246.6939, -245.7192, -244.7981, -243.9319, -243.1223, -242.3704, 
          -241.6776, -241.0452, -240.4744, -239.9661, -239.5216, -239.1417, 
          -238.8274, -238.5795, -238.3987, -238.2858, -238.2414, -238.2660, 
          -238.3601, -238.5240, -238.7582, -239.0629, -239.4382, -239.8843, 
          -240.4013, -240.9891, -241.6476, -242.3767, -243.1762, -244.0458, 
          -244.9853, -245.9941)
}
@

\begin{figure}[htbp]
\begin{center}
<<proflike,echo=TRUE,fig=TRUE>>=
plot(x,jj-max(jj),type='b',xlab="log-odds",ylab="support")
abline(h=c(0,-2))
@
\caption{Profile likelihood \label{proflike} for odds ratio
  $\log\left(D_s/D\right)$}
  \end{center}
\end{figure}
